---
title: "01.vectors"
author: "Giancarlo Prioli Cardozo"
format: html
editor: visual
---

#vectors

vectors are the most basic and fundamental data structure in R. Every element in a vector must have the same data type. Most operations in R are vectorized, meaning they are applied automatically to every element

##creating vectors

vectors are created with the `c()` function

```{r}
# Numeric vector
heart_rate <- c(72, 81, 68, 95, 88, 76, 70, 91, 85, 79, 74, 83)

# Character vector
patient_group <- c("A", "B", "A", "B")

# Logical vector
critical <- c(FALSE, FALSE, FALSE, TRUE)
```

##basic properties

these functions describe the structure of a vector

```{r}
class(heart_rate)
typeof(heart_rate)
length(heart_rate)
str(heart_rate)
```

##summary statistics

many statistical functions operate directly on numeric vectors

```{r}
mean(heart_rate)
median(heart_rate)
sd(heart_rate)
range(heart_rate)
summary(heart_rate)
```

##vectorized operations

arithmetic operations are automatically applied to every element

```{r}
heart_rate + 5
heart_rate * 1.10
heart_rate - mean(heart_rate)
```

##logical comparisons

logical expressions return a vector of `TRUE` and `FALSE`

```{r}
heart_rate > 80

oxygen_saturation <- c(
  98, 96, 99, 92, 95, 97,
  99, 91, 94, 96, 98, 95
)

oxygen_saturation < 95
```

##indexing

selecting elements by position

```{r}
heart_rate[1]

heart_rate[c(2, 5, 8)]

heart_rate[-1]
```

##filtering

logical vectors can be used to filter observations

```{r}
heart_rate[heart_rate > 80]

oxygen_saturation[oxygen_saturation >= 97]

heart_rate[
  heart_rate > 80 &
  oxygen_saturation < 95
]
```

##creating sequences

useful functions for generating vectors

```{r}
1:12

seq(0, 20, by = 5)

rep(c("Day", "Night"), times = 6)
```

##named vectors

names can be assigned to improve readability

```{r}
names(heart_rate) <- paste0("P", 1:12)

heart_rate
```

##missing values

missing values are represented by `NA`

```{r}
heart_rate[4] <- NA

is.na(heart_rate)

sum(is.na(heart_rate))

mean(heart_rate, na.rm = TRUE)
```

missing values can be replaced

```{r}
heart_rate[is.na(heart_rate)] <-
  median(heart_rate, na.rm = TRUE)

heart_rate
```

##type coercion

all elements in a vector must share the same type. R automatically coerces values when necessary

```{r}
mixed_vector <- c(1, 2, "3")

mixed_vector

typeof(mixed_vector)

as.numeric(mixed_vector)

as.character(heart_rate)
```

##useful functions

commonly used functions when working with vectors

```{r}
which(heart_rate > 85)

sort(heart_rate)

unique(rep(c("Day", "Night"), times = 6))
```

##practice

a small manually created dataset

```{r}
temperature <- c(
  36.5, 36.8, 37.1, 38.2,
  37.8, 36.7, 36.9, 38.5,
  37.3, 36.6, 36.8, 37.0
)

alert <- heart_rate > 85 | temperature > 38

alert
```
