---
title: "Lists"
author: "Giancarlo Prioli Cardozo"
format: html
editor: visual
---

#lists

lists can store objects of different types. They are commonly used to represent datasets, model outputs, and more complex structures.

##creating lists

lists are created with the `list()` function

```{r}
#simplified patient record
patient <- list(
  id = 101,
  age = 64,
  sex = "Female",
  hypertension = TRUE,
  glucose = c(118, 121, 125, 119),
  medications = c("Losartan", "Metformin")
)
```

##basic properties

these functions describe the structure and components of a list

```{r}
str(patient)
names(patient)
length(patient)
```

##accessing elements

elements can be accessed using `$`, `[[]]`, or numerical indexing

```{r}
patient$age
patient[["sex"]]
patient[[5]]

#mean glucose level
mean(patient$glucose)
```

##modifying lists

adding, updating, and removing elements in a list

```{r}
# Adding information
patient$diagnosis <- "Type 2 Diabetes"

# Updating values
patient$age <- 65

# Removing an element
patient$hypertension <- NULL
```

##nested lists

lists can contain other lists

```{r}
#list of patients
patients <- list(
  list(id = 1, age = 45, glucose = 95),
  list(id = 2, age = 61, glucose = 132),
  list(id = 3, age = 54, glucose = 108)
)

#accessing nested values
patients[[2]]$glucose
```

##extracting attributes

using functions like `sapply()` to extract values across list items

```{r}
ages <- sapply(patients, function(x) x$age)
glucose_levels <- sapply(patients, function(x) x$glucose)

ages
glucose_levels

#summary statistics
mean(ages)
mean(glucose_levels)
```

##converting to data frame

converting lists into data frames for analysis

```{r}
data <- data.frame(
  id = sapply(patients, `[[`, "id"),
  age = ages,
  glucose = glucose_levels
)

data

#risk classification
data$risk <- ifelse(
  data$glucose >= 126,
  "High",
  "Low"
)

data
```

##complex lists

lists can store different types of objects, including data frames and summary metrics

```{r}
study <- list(
  data = data,
  mean_glucose = mean(data$glucose),
  n_patients = nrow(data),
  outcome = "glucose"
)

str(study)

#accessing information
study$mean_glucose
study$data
```