---
title: "functions"
author: "Giancarlo Prioli Cardozo"
format: html
editor: visual
---

#functions

functions are blocks of code that perform specific tasks. they help avoid repetition and make the analytical code more organized and readable

##built-in functions

R has many built-in functions for common calculations in health data

```{r}
weights_kg <- c(65, 72, 88, 54, 91)

mean(weights_kg)
median(weights_kg)
round(mean(weights_kg), 1)
```

##creating basic functions

you can create custom functions using the `function()` followed by the code block

```{r}
#Body Mass Index (BMI)
calculate_bmi <- function(weight, height) {
  bmi <- weight / (height^2)
  return(bmi)
}

calculate_bmi(weight = 70, height = 1.75)
```

##default arguments

you can set default values for arguments if they are not provided by the user

```{r}
#convert glucose from mmol/L to mg/dL
convert_glucose <- function(value, multiply_by = 18) {
  return(value * multiply_by)
}

convert_glucose(5.5) #default multiplier
convert_glucose(5.5, multiply_by = 18.018) #uses molar mass
```

##returning values

functions automatically return the last evaluated expression, but `return()` makes it explicit and can stop execution early

```{r}
assess_fever <- function(temp_c) {
  if (is.na(temp_c)) {
    return("Missing data")
  }
  
  if (temp_c > 38.0) {
    return("Fever")
  }
  
  "Normal"
}

assess_fever(37.5)
assess_fever(38.6)
```

##multiple returns

if you need to return multiple values, you must store them in a list or vector

```{r}
patient_summary <- function(sbp, dbp) {
  map <- dbp + (1/3) * (sbp - dbp)
  pulse_pressure <- sbp - dbp
  
  list(
    MAP = round(map, 1),
    Pulse_Pressure = pulse_pressure
  )
}

patient_summary(120, 80)
```

##scoping

variables created inside a function are local and do not affect the global code

```{r}
heart_rate <- 72

update_hr <- function(new_hr) {
  heart_rate <- new_hr
  return(heart_rate)
}

update_hr(90)

#global variable remains unchanged
heart_rate 
```

##anonymous functions

short functions created on the fly, often used with `apply()` or `sapply()`

```{r}
patient_ages <- list(p1 = 45, p2 = 60, p3 = 33)

#add 1y to every patient age
sapply(patient_ages, function(age) age + 1)
```
