---
title: "control_flow"
author: "Giancarlo Prioli Cardozo"
format: html
editor: visual
---

#control flow

control flow statements determine the order in which code is executed. they allow for decision making and iteration over clinical datasets.

##if statements

executes a specific block of code only if a given condition is true

```{r}
patient_temp <- 38.5

if (patient_temp > 38.0) {
  print("Alert: Patient has a fever.")
}
```

##if else statements

provides an alternative action if the initial condition is FALSE

```{r}
oxygen_sat <- 93

if (oxygen_sat >= 95) {
  print("Oxygen levels are normal.")
} else {
  print("Warning: Administer oxygen.")
}
```

##else if chains

allows checking multiple conditional thresholds back to back

```{r}
systolic_bp <- 145

if (systolic_bp < 120) {
  print("Normal")
} else if (systolic_bp < 130) {
  print("Elevated")
} else if (systolic_bp < 140) {
  print("Stage 1 Hypertension")
} else {
  print("Stage 2 Hypertension")
}
```

##vectorized ifelse

`ifelse()` is useful because it evaluates a condition across a whole vector at once

```{r}
glucose_levels <- c(95, 110, 135, 88, 126)

status <- ifelse(
  glucose_levels >= 126, 
  "Diabetes", 
  "Normal/Pre-diabetes"
)

status
```

##for loops

repeats a block of code for each element in a sequence, vector, or list

```{r}
patient_ids <- c("P01", "P02", "P03")

for (id in patient_ids) {
  message <- paste("Retrieving electronic health record for:", id)
  print(message)
}
```

##while loops

repeats code as long as a condition remains true. useful when the number of iterations is unknown

```{r}
days_in_hospital <- 1
recovery_score <- 40

while (recovery_score < 80) {
  recovery_score <- recovery_score + 15
  days_in_hospital <- days_in_hospital + 1
}

days_in_hospital
```

##break and next

`break` stops a loop completely, while `next` skips only the current iteration

```{r}
heart_rates <- c(70, 75, 150, 80, 85)

#using break for critical events
for (hr in heart_rates) {
  if (hr > 120) {
    print("Tachycardia detected. Stopping exercise test.")
    break
  }
  print(paste("Monitoring HR:", hr))
}

#using next to ignore artifacts
for (hr in heart_rates) {
  if (hr > 120) {
    next 
  }
  print(paste("Valid reading saved:", hr))
}
```